On Determinism of Game Engines used for Simulation-based Autonomous Vehicle Verification

### 2020-11-13 paper/cav_determinisim_main.tex
Added diagrams for variance description and plot for nice analysis
Added diagrams for variance description, updated clearer plots for pre post and added  plot for nice analysis
Corrected some spelling mistakes
GC added NICE

% ***************************************************
%  TITLE
% ***************************************************
\title{On Determinism of Game Engines used for Simulation-based Autonomous Vehicle Verification}
% \title{On the Precision of Game Engines used for Simulation-based Autonomous Vehicle Verification}
\author[1,3]{Abanoub Ghobrial\thanks{$^{1}${\footnotesize \{abanoub.ghobrial, greg.chance, kerstin.eder\}@bristol.ac.uk}}}
\author[1,3]{Greg Chance}
\author[2,3]{Severin Lemaignan\thanks{$^{2}${\footnotesize \{severin.lemaignan, tony.pipe\}@brl.ac.uk}}}
\author[2,3]{Tony Pipe}
\author[1,3]{Kerstin Eder}
\affil[1]{Trustworthy Systems Laboratory, University of Bristol, Bristol, UK}
\affil[2]{University of West of England, Bristol, UK}
\affil[3]{Bristol Robotics Laboratory, Bristol, UK}
\maketitle

@@

\begin{figure}[!t]
    \centering
    \begin{subfigure}{.48\textwidth}
        \includegraphics[width=1\textwidth]{Other/Figures/Variance_predicition_tolerance_definition_diagram_a.pdf}
        \caption{}
        \label{variance_description_a}
    \end{subfigure}

    \begin{subfigure}{.48\textwidth}
        \includegraphics[width=1\textwidth]{Other/Figures/Variance_predicition_tolerance_definition_diagram_b.pdf}
        \caption{}
        \label{variance_description_b}
    \end{subfigure}
    \caption{Demonstration of variance, precision, tolderance and determinism}
    \label{variance_description}
\end{figure}

@@

% ======= Sources of non-determinism
\subsection{Potential Sources of Non-Determinism} \label{s:nondeterminisimSources}

@@

% ======= Floating-Point Numbers
\noindent\underline{\textit{Floating-Point Arithmetic:}}

@@

% Translating to the usage of AV simulations in physics engines,
In the context of AV simulation, such rounding errors could result in precision issues of, for example, actor positions within the environment, leading to falsely satisfied or failed assertions.
% \todo[inline]{state what assertions are (used for) GC-see below} 
%In the field of AV verification, assertions are the immutable rules and regulations that describe safe and legal driving. 
%
Thus, one could jump to the conclusion that floating-point errors cause non-deterministic behaviour, resulting in loss of repeatability.

@@

\noindent\underline{\textit{Error Correcting Code (ECC) Memory:}}
ECC Memory is used ubiquitously in commercial simulation facilities and servers to detect and correct single bit errors in DRAM memory~\cite{Dell1997}. Single bit errors may occur due to malfunctioning hardware, ionising radiation (background cosmic or environmental sources) or from electromagnetic radiation~\cite{dodd2003basic}. If single bit errors go uncorrected then subsequent computational processing will produce incorrect results, potentially giving rise to non-determinism due to the probabilistic nature of such errors occurring. Estimating the rate of error is difficult and dependent on hardware, environment and computer cycles~\cite{mielke2008bit}.

\noindent\underline{\textit{Actor Navigation:}}
There is existing evidence to suggest actor navigation, typically pedestrians, could be the cause of non-deterministic simulation behaviour~\cite{CARLABenchmark}.
% 
Unreal Engine is the underlying framework for the CARLA simulator which uses the A* algorithm for actor navigation\footnote{\url{https://www.oreilly.com/library/view/unreal-engine-4/9781784393120/ch04s03.html} from~\cite{newton2016unreal}}.
%
The A* algorithm~\cite{AStarBook} will give deterministic outcomes as long as the environment is deterministic~\cite{AirsimUnrealArticle, UnrealAIDocumentation}.

Unless other sources that can alter the fixed environment exist, the A* algorithm should give deterministic results. Thus, actor navigation should not be considered a source of non-determinism.
%
% Therefore, the determinism of the actor path planning depends on the environment and the update to actor states within it and how navigation meshes are created or modelled, e.g. shape, granularity. 
%
% It is interesting to note that changes could occur to navigation meshes if they are dynamically loaded as the simulation proceeds which can reduce computational load for large area maps\footnote{\url{https://docs.unrealengine.com/en-US/BlueprintAPI/AI/Navigation/RegisterNavigationInvoker/}}.
% \todo[inline]{unclear :( GC- re-written}
\\\\
\noindent\underline{\textit{Game Engine Setup:}}
The type and version of the engine code executed should be considered, paying attention to the control of pseudo-random numbers, fixed physics calculation steps ($dt$), fixed actor navigation mesh, deterministic ego vehicle controllers and engine texture loading rates especially for perception stack testing. %
% \todo[inline]{introduce actor navigation mesh GC-section on actor navigation brought forward}
% \todo[inline]{introduce ego vehicle earlier on GC added line 99}
For example, in Unreal Editor the \textit{unit}\footnote{\url{https://docs.unrealengine.com/en-US/Engine/Performance/StatCommands/}} command can be used to monitor performance metrics such as \textit{Frame} which reports 
% to determine processing bottlenecks by indicating\todo[inline]{what does indicating mean? GC-displaying, returning or showing} 
the total time spent generating one frame, \textit{Game} for game loop execution time and \textit{Draw} for render thread time. 
%
Wrt.\ perception stack testing, weather and lighting conditions in the game engine should be controlled as well as any other dynamic elements to the simulation environment, e.g.\ reflections from surface water. 
\\\\
\noindent\underline{\textit{NUMA Non-Uniform Memory Access:}}
For a repeated test that operates over a number of cores based on a CPU scheduling policy, memory access time may vary depending on the physical memory location relative to the processor. Typically a core can access its own memory with lower latency than that of another core resulting in lower interprocessor data transfer cost~\cite{nieplocha1996global}. 
%
Changes in latency between repeated tests may, in the worst case, cause the game engine to operate non-deterministically, or, perhaps, simply with an increased data transfer cost, i.e.\ slower. 
%
By binding a process to a specific core for the duration of its execution, the variations in data transfer time can be minimised.
\\\\
\noindent\underline{\textit{Scheduling, Concurrency and Parallelisation:}}
% \todo[inline]{is this at runtime?} %
% info at:
% http://www.informit.com/articles/article.aspx?p=169479&seqNum=3
% https://www.os-book.com/OS8/os8c/slide-dir/PDF-dir/ch5.pdf
% https://en.wikipedia.org/wiki/Concurrency_(computer_science)
% https://en.wikipedia.org/wiki/Concurrent_computing
% https://en.wikipedia.org/wiki/Race_condition#cite_note-1
Runtime scheduling is a resource management method for sharing computational resources between tasks of different or equal priority where tasks are executed dependent on the scheduler policy of the operating system. A scheduler policy may be optimised in many ways such as for task throughput, deadline delivery or minimum latency~\cite{liu1973scheduling}. 
% \todo[inline]{is this the scheduling that is done by the operating system, i.e. at runtime? GC - runtime, added to 1st para.}
%
Changing the scheduling policy and thread priorities may increase simulation variance.

It is, however, unlikely that thread priorities or scheduling policy would occur during repeated controlled tests for the same hardware and operating system configuration. Given that hardware is considered deterministic and if the thread execution order does not change between tests, then for the same hardware and operating system configuration the same output should be given.

@@

\\\\

@@

\noindent\underline{\textit{Summary:}}
We have investigated the potential sources of non-determinism affecting game engines 
and explored the impact they may have on simulation variance. 
% \todo[inline,color=red!40]{GC need to decide if fp is a source or not: Assuming deterministic hardware, there is no contribution to simulation variance due to floating-point representation and any rounding errors due to bit width limits should be repeatable. }
Memory checking not withstanding, errors associated with the lack of ECC are likely to be minimal unless there is significant background radiation or 1000's of hours of computation are expected.
\todo[inline]{So, do commercial HPC facilities have ECC memory?}
\todo[inline,color=green!40]{GC yes see Dell1997 "Today all servers except the very low end support Single Error Correct (SEC) ECC"}
\todo[inline,color=red!40]{old: To ensure repeatable simulation outcomes the physics setting $dt$ must be fixed, along with any actor navigation meshes, random number seeds, game engine setup and simulation specific parameters.}
To ensure precise simulation outcomes the physics setting $dt$ must be fixed, along with any actor navigation meshes, random number seeds, game engine setup and simulation specific parameters.
\todo[inline,color=green!40]{GC alternative: To ensure precise simulation outcomes the physics setting $dt$ must be fixed, along with any actor navigation meshes, random number seeds, game engine setup and simulation specific parameters.}
NUMA should only affect interprocessor data transfer cost and without control measures will only make the computation cycle longer. Relative access times between different caches are likely to be small although may have a more pronounced impact on high throughput systems, e.g.\ HPC. 
Basic thread scheduling should not affect the simulation's determinism unless changing scheduling policy, operating system or migrating between machines with different setups. 
However, should new and unexpected threads start during the simulation, then the interruption to execution order or additional resource demand may affect timing of subsequent steps, thus reducing the number of physics updates within a game loop. Likewise, uncontrolled allocation of hardware resources such as CPUs or GPUs can potentially give rise to non-determinism. 

@@

\begin{figure*}[t]
    \centering
    \begin{subfigure}{.49\textwidth}
        \includegraphics[width=1\textwidth]{Other/Figures/CarsCollisionCG25_V4.pdf}
        \caption{}
        \label{CarsCollisionCG25}
    \end{subfigure}
    \begin{subfigure}{.49\textwidth}
        \includegraphics[width=1\textwidth]{Other/Figures/CarsCollisionPrePostV3.pdf}
        \caption{}
        \label{CarsCollisionPrePost}
    \end{subfigure}
    \caption{Vehicle to vehicle collision (test 2) showing (a) mean deviation against simulation time for 25\% resource utilisation and (b) mean deviation pre- and post-collision against resource utilisation. The simulation noise floor is shown in (a) which is the empirical lower limit of deviation for the hardware reported in this study.}
\end{figure*}

@@

\begin{figure*}[h]
    \centering
    \begin{subfigure}{.49\textwidth}
        \includegraphics[width=1\textwidth]{Other/Figures/CarsPeopleCollsionsCG25_V3.pdf}
        \caption{}
        \label{CarsPeopleCollsionsCG25}
    \end{subfigure}
    \begin{subfigure}{.49\textwidth}
        \includegraphics[width=1\textwidth]{Other/Figures/CarsPeopleCollisionPrePostV3.pdf}
        \caption{}
        \label{CarsPeopleCollisionPrePost}
    \end{subfigure}
    \caption{Vehicle to pedestrian collision (test 4) showing (a) maximum deviation against simulation time for 25\% stress and (b) maximum deviation pre- and post-collision for different resource utilisation levels.}
\end{figure*}

@@

>>>>

% ***************************************************
%  TITLE
% ***************************************************
\title{On Determinism of Game Engines used for Simulation-based Autonomous Vehicle Verification}
% \title{On the Precision of Game Engines used for Simulation-based Autonomous Vehicle Verification}
\author[1,3]{Abanoub Ghobrial\thanks{$^{1}${\footnotesize \{abanoub.ghobrial, greg.chance, kerstin.eder\}@bristol.ac.uk}}}
\author[1,3]{Greg Chance}
\author[2,3]{S\'everin Lemaignan\thanks{$^{2}${\footnotesize \{severin.lemaignan, tony.pipe\}@brl.ac.uk}}}
\author[2,3]{Tony Pipe}
\author[1,3]{Kerstin Eder}
\affil[1]{Trustworthy Systems Laboratory, University of Bristol, Bristol, UK}
\affil[2]{University of West of England, Bristol, UK}
\affil[3]{Bristol Robotics Laboratory, Bristol, UK}
\maketitle

@@

\begin{figure}[!t]
    \centering
    \begin{subfigure}{.48\textwidth}
        \includegraphics[width=1\textwidth]{Other/Figures/Variance_predicition_tolerance_definition_diagram_a.pdf}
        \caption{Non-deterministic}
        \label{variance_description_a}
    \end{subfigure}

    \begin{subfigure}{.48\textwidth}
        \includegraphics[width=1\textwidth]{Other/Figures/Variance_predicition_tolerance_definition_diagram_b.pdf}
        \caption{Deterministic}
        \label{variance_description_b}
    \end{subfigure}
    \caption{Demonstration of variance, precision, tolerance and determinism}
    \label{variance_description}
\end{figure}

@@

% ======= Sources of non-determinism
\section{Potential Sources of Non-Determinism} \label{s:nondeterminisimSources}

@@

% ======= Floating-Point Numbers
\subsection{Floating-Point Arithmetic}

@@

% Translating to the usage of AV simulations in physics engines,
In the context of AV simulation, such rounding errors could result in accuracy issues of, for example, actor positions within the environment, leading to falsely satisfied or failed assertions.
% \todo[inline]{state what assertions are (used for) GC-see below} 
%In the field of AV verification, assertions are the immutable rules and regulations that describe safe and legal driving. 
%
Thus, one could jump to the conclusion that floating-point errors cause non-deterministic behaviour, resulting in loss of repeatability.

@@

\subsection{Scheduling, Concurrency and Parallelisation}
% \todo[inline]{is this at runtime?} %
% info at:
% http://www.informit.com/articles/article.aspx?p=169479&seqNum=3
% https://www.os-book.com/OS8/os8c/slide-dir/PDF-dir/ch5.pdf
% https://en.wikipedia.org/wiki/Concurrency_(computer_science)
% https://en.wikipedia.org/wiki/Concurrent_computing
% https://en.wikipedia.org/wiki/Race_condition#cite_note-1
Runtime scheduling is a resource management method for sharing computational resources between tasks of different or equal priority where tasks are executed dependent on the scheduler policy of the operating system. A scheduler policy may be optimised in many ways such as for task throughput, deadline delivery or minimum latency~\cite{liu1973scheduling}. 
% \todo[inline]{is this the scheduling that is done by the operating system, i.e. at runtime? GC - runtime, added to 1st para.}
%
Changing the scheduling policy and thread priorities may increase simulation variance.

It is, however, unlikely that changes to thread priorities or scheduling policy would occur during repeated controlled tests for the same hardware and operating system configuration. Given that hardware is considered deterministic and if the thread execution order does not change between tests, then for the same hardware and operating system configuration the same output should be given.

@@

@@

\subsection{NUMA Non-Uniform Memory Access}
For a repeated test that operates over a number of cores based on a CPU scheduling policy, memory access time may vary depending on the physical memory location relative to the processor. Typically a core can access its own memory with lower latency than that of another core resulting in lower interprocessor data transfer cost~\cite{nieplocha1996global}. 
%
Changes in latency between repeated tests may, in the worst case, cause the game engine to operate non-deterministically if tasks are processed out of sequence using equal priority scheduling, or, perhaps, simply with an increased data transfer cost, i.e.\ slower. 
%
By binding a process to a specific core for the duration of its execution, the variations in data transfer time can be minimised.

\subsection{Error Correcting Code (ECC) Memory}
ECC Memory is used ubiquitously in commercial simulation facilities and servers to detect and correct single bit errors in DRAM memory~\cite{Dell1997}. Single bit errors may occur due to malfunctioning hardware, ionising radiation (background cosmic or environmental sources) or from electromagnetic radiation~\cite{dodd2003basic}. If single bit errors go uncorrected then subsequent computational processing will produce incorrect results, potentially giving rise to non-determinism due to the probabilistic nature of such errors occurring. Estimating the rate of error is difficult and dependent on hardware, environment and computer cycles~\cite{mielke2008bit}.

\subsection{Game Engine Setup}
The type and version of the engine code executed should be considered, paying attention to the control of pseudo-random numbers, fixed physics calculation steps ($dt$), fixed actor navigation mesh, deterministic ego vehicle controllers and engine texture loading rates especially for perception stack testing. %
%
% It must be noted that if $dt$ cannot be fixed then deterministic results are unlikely.
% \todo[inline]{introduce actor navigation mesh GC-section on actor navigation brought forward}
% \todo[inline]{introduce ego vehicle earlier on GC added line 99}
For example, in Unreal Editor the \textit{unit}\footnote{\url{https://docs.unrealengine.com/en-US/Engine/Performance/StatCommands/}} command can be used to monitor performance metrics such as \textit{Frame} which reports 
% to determine processing bottlenecks by indicating\todo[inline]{what does indicating mean? GC-displaying, returning or showing} 
the total time spent generating one frame, \textit{Game} for game loop execution time and \textit{Draw} for render thread time. 
%
With respect to perception stack testing, weather and lighting conditions in the game engine should be controlled as well as any other dynamic elements to the simulation environment, e.g.\ reflections from surface water, ensuring textures are not randomly generated.

\subsection{Actor Navigation}
There is existing evidence to suggest actor navigation, typically pedestrians, could be the cause of non-deterministic simulation behaviour~\cite{CARLABenchmark}.
% 
Unreal Engine is the underlying framework for the CARLA simulator which uses the A* algorithm for actor navigation\footnote{\url{https://www.oreilly.com/library/view/unreal-engine-4/9781784393120/ch04s03.html} from~\cite{newton2016unreal}}.
%
The A* algorithm~\cite{AStarBook} will give deterministic outcomes as long as the environment is deterministic~\cite{AirsimUnrealArticle, UnrealAIDocumentation}.

Unless other sources that can alter the fixed environment exist, the A* algorithm should give deterministic results. Thus, actor navigation should not be considered a source of non-determinism.
%
% Therefore, the determinism of the actor path planning depends on the environment and the update to actor states within it and how navigation meshes are created or modelled, e.g. shape, granularity. 
%
% It is interesting to note that changes could occur to navigation meshes if they are dynamically loaded as the simulation proceeds which can reduce computational load for large area maps\footnote{\url{https://docs.unrealengine.com/en-US/BlueprintAPI/AI/Navigation/RegisterNavigationInvoker/}}.
% \todo[inline]{unclear :( GC- re-written}

\subsection{Summary}
We have investigated the potential sources of non-determinism affecting game engines 
and explored the impact they may have on simulation variance. 
%
% \todo[inline,color=red!40]{GC need to decide if fp is a source or not: Assuming deterministic hardware, there is no contribution to simulation variance due to floating-point representation and any rounding errors due to bit width limits should be repeatable. }
%
Memory checking not withstanding, errors associated with the lack of ECC are likely to be minimal unless there is significant background radiation or 1000's of hours of computation are expected.
\todo[inline]{So, do commercial HPC facilities have ECC memory?}
\todo[inline,color=green!40]{GC yes see Dell1997 "Today all servers except the very low end support Single Error Correct (SEC) ECC"}
%
\todo[inline,color=red!40]{old: To ensure repeatable simulation outcomes the physics setting $dt$ must be fixed, along with any actor navigation meshes, random number seeds, game engine setup and simulation specific parameters.}
To ensure precise simulation outcomes the physics setting $dt$ must be fixed, along with any actor navigation meshes, random number seeds, game engine setup and simulation specific parameters.
\todo[inline,color=green!40]{GC alternative: To ensure precise simulation outcomes the physics setting $dt$ must be fixed, along with any actor navigation meshes, random number seeds, game engine setup and simulation specific parameters.}
%
NUMA should only affect interprocessor data transfer cost and without control measures will only make the computation cycle longer. Relative access times between different caches are likely to be small although may have a more pronounced impact on high throughput systems, e.g.\ HPC. 
%
Basic thread scheduling should not affect the simulation's determinism unless changing scheduling policy, operating system or migrating between machines with different setups. 
%
However, should new and unexpected threads start during the simulation, then the interruption to execution order or additional resource demand may affect timing of subsequent steps, thus reducing the number of physics updates within a game loop. Likewise, uncontrolled allocation of hardware resources such as CPUs or GPUs can potentially give rise to non-determinism. 
%

@@

\begin{figure*}[t]
    \centering
    \begin{subfigure}{.49\textwidth}
        \includegraphics[width=1\textwidth]{Other/Figures/CarsCollisionCG25_V4.pdf}
        \caption{}
        \label{CarsCollisionCG25}
    \end{subfigure}
    \begin{subfigure}{.49\textwidth}
        \includegraphics[width=1\textwidth]{Other/Figures/CarsCollisionPrePostV4.pdf}
        \caption{}
        \label{CarsCollisionPrePost}
    \end{subfigure}
    \caption{Vehicle to vehicle collision (test 2) showing (a) mean deviation against simulation time for 25\% resource utilisation and (b) mean deviation pre- and post-collision against resource utilisation. The simulation noise floor is shown in (a) which is the empirical lower limit of deviation for the hardware reported in this study.}
\end{figure*}

@@

\begin{figure*}[h]
    \centering
    \begin{subfigure}{.49\textwidth}
        \includegraphics[width=1\textwidth]{Other/Figures/CarsPeopleCollsionsCG25_V3.pdf}
        \caption{}
        \label{CarsPeopleCollsionsCG25}
    \end{subfigure}
    \begin{subfigure}{.49\textwidth}
        \includegraphics[width=1\textwidth]{Other/Figures/CarsPeopleCollisionPrePostV4.pdf}
        \caption{}
        \label{CarsPeopleCollisionPrePost}
    \end{subfigure}
    \caption{Vehicle to pedestrian collision (test 4) showing (a) maximum deviation against simulation time for 25\% stress and (b) maximum deviation pre- and post-collision for different resource utilisation levels.}
\end{figure*}

@@

\subsection{Process Scheduling Priority}
An investigation into the response of process scheduling on the simulation variance was undertaken. The experiment was repeated using test 1 but altering the process scheduling priority using the Ubuntu program \texttt{NICE}. Setting a higher priority for the simulator process with respect to the resource utilisation processes, it was possible to determine if scheduling could account for the increased simulation variance when the system is under stress.
%
To give a process a high priority a negative NICE value is set with the lowest being -20. To decrease the priority a positive value is set, up to +19. The default NICE value is 0.

The results presented in a box and whisker format, Fig.~\ref{NICEExperimentStressSummary}, show that decreasing the priority (right hand side of plot) has little effect on simulation variance when compared to a 0 NICE value (central plot bars). Increasing priority (left hand side of plot) reduced variance for the 75\% resource utilisation  but this did not account for all the difference in the observed results. 
%
This unaccounted difference in the variance may be due to the lack of  control that \texttt{NICE} has over the processes~\cite{Abanoub can you cite that quote you found here please}.
